---
title: "Capstone Project: Fixed Effects Model"
author: Megan Marsh
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  html:
    embed-resources: true
editor: source
---


```{r}
library(plm)
CapstoneProjectUpdated2Clean <- read.csv("Capstone/CapstoneProjectUpdated2 - CapstoneProjectUpdated (1).csv")  
panel_data <- pdata.frame(CapstoneProjectUpdated2Clean, index = c("UnitID", "Year"))
```


```{r}
#fixed effects model (individual effects within)
fixed_effects_model <- plm(GraduationRate ~ CovidIndicator + PercentWhite + PercentWomen + 
                               Staff:CovidIndicator + AvgInStateTuition + RetentionRate + 
                               TotalEnrollment + AvgPellGrant + AvgStudentLoans, 
                             data = panel_data, model = "within")
```

```{r}
summary(fixed_effects_model)

#random effects model
random_effects_model <- plm(GraduationRate ~ CovidIndicator + PercentWhite + PercentWomen +
                               Staff:CovidIndicator + AvgInStateTuition + RetentionRate +
                             TotalEnrollment + AvgPellGrant + AvgStudentLoans,
                            data = panel_data, model = "random")
summary(random_effects_model)
```


```{r}
#hausman test: to decide if between fixed and random effects
phtest(fixed_effects_model, random_effects_model)
#fixed Effects more appropriate
```

The Hausman Test verified the use of the fixed-effects model over the random-effects model. 

```{r}
#breusch-godfrey test: to detect serial correlation
pbgtest(fixed_effects_model)
```

Based on the Breusch-Godfrey/Wooldridge test, there is no statistically significant evidence of serial correlation in the idiosyncratic errors. This suggests that the error terms for a given institution across different time periods are likely independent after accounting for the other variables in your model.

```{r}
#breusch-pagan test for cross-sectional dependence 
#null hypothesis: no cross-sectional dependence in the error terms (or residuals) of the different cross-sectional units
pcdtest(fixed_effects_model, test = c("lm")) 
```

The Breusch-Pagan Test yields a very small p-values (below the chosen significance level of 0.05), providing sufficient evidence to reject the null hypothesis of no cross-sectional dependence in the error terms (or residuals) of the different cross-sectional units. This mean robust standard errors must be used on model to produce reliable results. 

```{r}
library(plm)
library(sandwich)
library(lmtest)

robust_se <- coeftest(fixed_effects_model, vcov. = vcovHC(fixed_effects_model, type = "HC1", cluster = "group"))
print(robust_se)
```


```{r}
library(ggplot2)

predicted_values2 <- predict(fixed_effects_model, newdata = panel_data)

ggplot(data.frame(CapstoneProjectUpdated2Clean$GraduationRate, predicted_values2), 
       aes(x = CapstoneProjectUpdated2Clean$GraduationRate, y = predicted_values2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(title = "Predicted vs. Actual Graduation Rates",
       x = "Actual Graduation Rate",
       y = "Predicted Graduation Rate")
```